Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-09-30 astronum_2018/astronum-xrb.tex
another pass

Another way we can reduce the number of zones for given accuracy is to use a more
accurate hydrodynamic method.  We are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions. \MarginPar{need some refs here} Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames. For smooth wave solutions we may expect to need
around half the number of zones per dimension, and since computational work
scales like $(\Delta x)^4$ in 3D, this could lower costs by a factor of around
10.
%
% This calculation is based on the number of points per wavelength required,
% to resolve a periodic wave, which (I believe) goes from 12 (2nd order) to
% about 7 (4th order). However, this is very idealized, and I'm not sure that
% it carries over to the reactions, so this may well be optimistic.
%

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{sdc_plot}
\caption{\label{fig:sdc} Helium mass fraction as a function of time
  over two timesteps for a detonation test problem with a 19-isotope
  reaction network.  The points represent individual calls to the
  righthand side function of our reaction network (see
  Eq. \ref{eq:syst}) and the density of points indicates how hard the
  integrator is working.  For the Strang case, we see that the first
  half of the reactions brings us off of the smooth solution
  (represented by the SDC curve), then the advection overcorrects,
  resulting in a discontinuous jump, and finally the second half of
  the reactions works to bring the state back to the solution curve.}
\end{figure}

@@

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of XRBs, one possibility for
multiscale modelling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a low Mach number or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the 2D shallow water model
would be negligible compared to those of the low Mach/compressible model, the overall cost of
this multiscale model would be determined by the size of the low Mach/compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
As with the more aggressive gridding strategy above, the hybrid approach
reduces the number of zones with the largest computational cost through mesh
refinement, and also reduces the computational costs where possible by
employing cheaper physical models in regions where this is possible.
An example of coupling compressible and low Mach number models implemented using the \amrex~framework, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implemenation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.

@@

>>>>

Another way we can reduce the number of zones for given accuracy is to use a more
accurate hydrodynamic method.  We are developing fully
fourth-order (in space and time) coupling of hydrodynamics and
reactions using the method of spectral deferred corrections (SDC) in
\castro, instead of the more commonly employed Strang splitting~\cite{strang:1968}.  This
provides us with two benefits.  First, the improved coupling actually
reduces the stiffness of the reactions, requiring fewer righthand side
calls and therefore reducing the expense of the
reactions.  Second, by moving to fourth
order, we may also be able to reduce our resolution requirements
needed for converged flames. For smooth wave solutions we may expect to need
around half the number of zones per dimension, and since computational work
scales like $(\Delta x)^4$ in 3D, this could lower costs by a factor of around
10.
%
% This calculation is based on the number of points per wavelength required,
% to resolve a periodic wave, which (I believe) goes from 12 (2nd order) to
% about 7 (4th order). However, this is very idealized, and I'm not sure that
% it carries over to the reactions, so this may well be optimistic.
%

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{sdc_plot}
\caption{\label{fig:sdc} Helium mass fraction as a function of time
  over two timesteps for a detonation test problem with a 19-isotope
  reaction network.  The points represent individual calls to the
  righthand side function of our reaction network (see
  Eq.~\ref{eq:syst}) and the density of points indicates how hard the
  integrator is working.  For the Strang case, we see that the first
  half of the reactions brings us off of the smooth solution
  (represented by the SDC curve), then the advection overcorrects,
  resulting in a discontinuous jump, and finally the second half of
  the reactions works to bring the state back to the solution curve.}
\end{figure}

@@

All of the above developments still involve fully compressible
hydrodynamics, but multiscale methods~\cite{weinan2011principles} where we couple different hydrodynamics
solvers together in a single simulation may ultimately be needed to model the full star while resolving the burning front.
%\MarginPar{Alice and Ian contribute here}
Such techniques rely on it being possible to decompose the dynamics of the system
into two scales, e.g.~a short and a long lengthscale, and that the processes at each
scale are at most weakly coupled with each other. In the context of XRBs, one possibility for
multiscale modeling would be to model the large scale dynamics using a shallow water model
(like that of~\cite{SPIT_ETAL02}) to capture the effects of the Coriolis force, and use
a low Mach number or
a fully compressible model in the vicinity of the burning front to capture
the turbulent burning processes. Given that the computational costs of the shallow water model
would be negligible compared to those of the low Mach/compressible model, the overall cost of
this multiscale model would be determined by the size of the low Mach/compressible region only. This
approach would therefore allow us to capture dynamics across the full star
without sacrificing resolution around the flame.
As with the more aggressive gridding strategy above, the hybrid approach
reduces the number of zones with the largest computational cost through mesh
refinement, and also reduces the computational costs where possible by
employing cheaper physical models in regions where this is possible.
An example of coupling compressible and low Mach number models implemented using the \amrex~framework, but in a terrestrial context,
is~\cite{Motheau2018}. An initial implementation, \cite{Harpole2018}, in an
astrophysical context, coupling shallow water to compressible models in both
Newtonian and relativistic gravity, indicates that considerable computational
efficiencies are possible, but that close attention to the coupling between the
models is needed.

@@

XRBs are multiscale, multiphysics problems that are challenging to
model.  Nevertheless, significant progress has been made in
multidimensional models of these events, through a set of
complementary algorithmic approaches.  Future advances in computer
architectures and algorithms will allow for even more realistic models
of these events, and help us connect to observations, ultimately telling
us about the neutron star itself.
